Type-Safe Modular Parsing

### 2017-01-03 paper/sections/CaseStudy.tex
Changes

To demonstrate the utility of our modular parsing approach, we
implemented parsers of the first 18 calculi from book \textit{Types
  and Programming Languages} (TAPL) []. TAPL introduces several
calculi from simple to complex, by gradually adding new features to
syntax. We compared our implementation with a non-modular implementation we found online [], which is also written in Scala and uses the same Packrat parsing library. The comparison is made from two aspects, source lines of code (SLOC) and execution time. \huang{done}\bruno{mention that you compare against an existing
  non-modular implementation.}

@@

We compared our implementation of the parsers with an implementation
by Ilya Klyuchnikov, available online []. Ilya's
implementation is suitable for comparison, because it is also
written in Scala using the same parser combinator library.
Furthermore, it includes parsers of all the 18 calculi we have, but
written in a non-modular way. Thus it is not able to reuse existing
code when those calculi share common features.

>>>>

To demonstrate the utility of our modular parsing approach, we
implemented parsers of the first 18 calculi from book \textit{Types
  and Programming Languages} (TAPL) \cite{pierce2002types}. TAPL introduces several
calculi from simple to complex, by gradually adding new features to
syntax. We compared our implementation with a non-modular implementation
we found online, which is also written in Scala and uses the same Packrat parsing library.
We counted source lines of code (SLOC) and measured execution time for both implementations.
Our implementation saves almost 70\% code comparing with that non-modular one.

\huang{done}\bruno{mention that you compare against an existing non-modular implementation.}

@@

We compared our implementation of the parsers with an implementation
by Ilya Klyuchnikov, available online\footnote{https://github.com/ilya-klyuchnikov/tapl-scala/}.
Ilya's implementation is suitable for comparison, because it is also
written in Scala using the same parser combinator library.
Furthermore, it includes parsers of all the 18 calculi we have, but
written in a non-modular way. Thus it is not able to reuse existing
code when those calculi share common features.

### 2017-01-03 paper/sections/PackratParsing.tex
Revised S4.
Changes

\paragraph{Longest Match Composition} Another problematic issue
in parser combinator libraries is the need for manually ordering
alternatives in a grammar.
Consider the grammar:
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <int> `+' <expr>
\end{grammar}
In Parsec, for instance, the parser
\begin{lstlisting}[language=PlainCode]
parser = parseInt <|> parseAdd
\end{lstlisting}
will only parse the input \inlinecode{"1 + 2"} to \inlinecode{"1"}, as \inlinecode{parseInt} successfully parses \inlinecode{"1"}
and terminates parsing. With traditional alternative
composition in recursive descent parsers, the parser in the front successfully parses a substring of the input, then it unexpectedly
finishes parsing and discards the rest, in spite that latter parsers might be abl e to parse the whole input.
%\huang{I've rewritten this paragraph}\bruno{The example is good, I think but the explanation is not.
%You want to say that when you know
%  the full grammar, you can figure out where ``try'' is
%  needed. Without the full grammar you'd need assume the worst
%  case. Make an effort to make your explanation cristal clear!
%I think you want to miss the abstract explanation that you give first,
%with the explanation about the concrete example.
%Start with ``The need for backtracking is also problematic
%in a modular setting. For example, suppose ...'' and synchronize
%the abstract explanation and the explanation for the example.
%}
In contrast, \inlinecode{parser = parseAdd <|> parseInt} works as expected.
In this case reordering the components in the alternatives ensures that
the \emph{longest match} for a string is parsed. Yet manual composition for longest match is inconvenient, and worst still, it is essentially non-modular. When the grammar is extended with new rules, programmers are supposed to \emph{manually} adjust
the order of parsers, which is requires rewritting previously written code.

@@

\noindent Unfortunatelly, this results in the worst exponential time complexity!

\subsection{Packrat Parsing}\label{subsec:packratparsing}
Fortunately some more advanced parsing techniques such as Packrat parsing~\cite{Ford2002}, address the limitations of simple parser combinators such as Parsec. Packrat parsers use memoization to record the result of applying each parser at each position of the input, so that repeated computation is eliminated.  Moreover, theoretically the algorithm behind Packrat parsers supports both direct and indirect left-recursion~\cite{warth2008}. All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique in our modular parsing library.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
In the current version of Scala (2.11), Packrat Parsing is included
in the Scala standard parser combinator library. The library provides a number of
parser combinators, together with some basic parsers. Below we present an example to illustrate Scala parsers.
%
%\bruno{So, we make a big fuss identifying the problems with other
%parser combinators in the previous subsection and now we present
%an example that does not suffer from any of the previous problems!
%What makes sense here is an example that uses: a \emph{left-recursive grammar}; the \emph{longest match combinator}; and where we can argue that
%there's no need for the user to do manual backtracking. Just use
%the example with the left-recursive grammar for parsing integers and additions and argue that none of the limitations
%discussed previously applies!
%}

\paragraph{Parsing a simple Arithmetic Language}
Suppose we want to parse a simple language with literals and
additions. The concrete syntax is as follows:
%\bruno{Here's a good opportunity to both improve the paper and make it
%more compact! This is the example that should be illustrated and
%discussed at the end of Section 2. Then we can just create a figure to
%refer to the code in Section 2 and refer again to the figure here. No
%need for two different examples.}

@@

@@

In the trait \lstinline{BaseParser}, \lstinline{lexical} is used for lexing. \lstinline{pLit} can parse an integer for \lstinline{Lit},
whereas \lstinline{pAdd} handles addition and creates an object of \lstinline{Add}. It parses two sub-expressions by calling \lstinline{pExpr}
recursively. Finally \lstinline{pExpr} composes \lstinline{pLit} and \lstinline{pAdd} using the alternative combinator with longest match.
Table~\ref{tab:packrat} gives a closer look at part of the API that we have used in our work.

>>>>

\paragraph{Longest Match Composition} Another problematic issue
in parser combinator libraries is the need for manually ordering
alternatives in a grammar.
Consider the grammar:
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <int> `+' <expr>
\end{grammar}
In Parsec, for instance, the parser
\begin{lstlisting}[language=PlainCode]
parser = parseInt <|> parseAdd
\end{lstlisting}
will only parse the input \inlinecode{"1 + 2"} to \inlinecode{"1"}, as \inlinecode{parseInt} successfully parses \inlinecode{"1"}
and terminates parsing. With traditional alternative
composition in recursive descent parsers, the parser in the front successfully parses a substring of the input, then it unexpectedly
finishes parsing and discards the rest, in spite that latter parsers might be able to parse the whole input.
%\huang{I've rewritten this paragraph}\bruno{The example is good, I think but the explanation is not.
%You want to say that when you know
%  the full grammar, you can figure out where ``try'' is
%  needed. Without the full grammar you'd need assume the worst
%  case. Make an effort to make your explanation cristal clear!
%I think you want to miss the abstract explanation that you give first,
%with the explanation about the concrete example.
%Start with ``The need for backtracking is also problematic
%in a modular setting. For example, suppose ...'' and synchronize
%the abstract explanation and the explanation for the example.
%}
In contrast, \inlinecode{parser = parseAdd <|> parseInt} works as expected.
In this case reordering the components in the alternatives ensures that
the \emph{longest match} for a string is parsed. Yet manual composition for longest match is inconvenient, and worst still, it is essentially non-modular. When the grammar is extended with new rules, programmers are supposed to \emph{manually} adjust
the order of parsers, which is requires rewritting previously written code.

@@

\noindent However, this results in the worst case exponential time complexity!

\subsection{Packrat Parsing}\label{subsec:packratparsing}
Fortunately some more advanced parsing techniques such as Packrat parsing~\cite{Ford2002}, address the limitations of simple parser combinators such as Parsec. Packrat parsers use memoization to record the result of applying each parser at each position of the input, so that repeated computation is eliminated.  Moreover, theoretically the algorithm behind Packrat parsers supports both direct and indirect left-recursion~\cite{warth2008}. All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique in our modular parsing library.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
Scala has a standard parser combinator library\footnote{https://github.com/scala/scala-parser-combinators}
which implements Packrat parsers.
The library provides a number of parser combinators, together with some basic parsers.
Below we present an example to illustrate Scala parsers.
%
%\bruno{So, we make a big fuss identifying the problems with other
%parser combinators in the previous subsection and now we present
%an example that does not suffer from any of the previous problems!
%What makes sense here is an example that uses: a \emph{left-recursive grammar}; the \emph{longest match combinator}; and where we can argue that
%there's no need for the user to do manual backtracking. Just use
%the example with the left-recursive grammar for parsing integers and additions and argue that none of the limitations
%discussed previously applies!
%}

\paragraph{Parsing a Simple Arithmetic Language}
Suppose we want to parse a simple language with literals and
additions. The concrete syntax is as follows:
%\bruno{Here's a good opportunity to both improve the paper and make it
%more compact! This is the example that should be illustrated and
%discussed at the end of Section 2. Then we can just create a figure to
%refer to the code in Section 2 and refer again to the figure here. No
%need for two different examples.}

@@

\huang{doesn't compile}

@@

In the trait \lstinline{ExprParser}, \lstinline{lexical} is used for lexing. \lstinline{pLit} can parse an integer for \lstinline{Lit},
whereas \lstinline{pAdd} handles addition and creates an object of \lstinline{Add}. It parses two sub-expressions by calling \lstinline{pExpr}
recursively. Finally \lstinline{pExpr} composes \lstinline{pLit} and \lstinline{pAdd} using the alternative combinator with longest match.
Table~\ref{tab:packrat} gives a closer look at part of the API that we have used in our work.
